Manuscript draft

### 2024-02-29 docs/manuscript.qmd
moved manuscript and other docs to separate folder

@@

An individual's social centrality may be determined by many factors. Differences in social position have been attributed to individual phenotypic characteristics, such as age, sex, or personality \[REFS\]. Of course, how an animal uses space is also an important driver of how many conspecifics it interacts with, and how frequently. The relationship between movement, space use, and social interaction patterns has been an area of active research in recent years. Webber et al [-@webber2023] have characterized the variation in patterns of animal movement and space use as the "spatial phenotype," as part of a larger framework for understanding the relationships between social and spatial drivers of animal behavior. In this light, the "spatial phenotype" may be considered in addition to more conventional phenotypic measures to explain why some individuals are more central than others in their social networks.

@@

-   Unusual GPS tracking coverage of the population–high proportion of individuals tracked, so we can get a population-level sense of movement behavior and social interaction patterns

    -   Communal roosts: many individuals together at once, variability in size/density of roosts. Used for information exchange \[REFS\] and also important for disease transmission \[or not–cf Elvira Nili\]. Breeding/nesting also occurs at/near roost sites, on cliffs.

# Research Questions

@@

@@

## Tagging and aging

@@

We subsetted the dataset to exclude periods of time that vultures spent in capture cages, in the wildlife hospital following poisonings or injuries, or in captivity prior to wild release (in the case of captive-born birds). We cleaned the data to remove invalid GPS fixes. Then we split data from each into three seasons. The fall, or "pre-breeding," season includes dates from 15 September through 14 December; the breeding season lasts from 15 December through 14 May of the following year; and the summer lasts from 15 May through 14 September. These season boundaries were chosen based on biological knowledge of the vultures' breeding timelines (see more details in the Study System section). Therefore, the analysis contained a total of 8 seasons, from the fall of 2020 through the 2023 breeding season (Table X). All of these data was reserved for use in proximity-based social network analysis. Since the GPS tags give few to no fixes overnight, and since reception is often very poor on the cliffs where vultures roost, we combined information from individuals' last locations of the evening and first locations of the morning to assign a roost site for each individual on each night \[more details\]. Full details about the data cleaning and roost assignment procedure can be found in the supplementary material.

@@

For each season, we calculated movement measures only for individuals tracked for at least 30 days (after removing low-point days as described above). We calculated flight altitude by extracting ground elevation information from elevation rasters obtained from the `elevatr` R package [@hollister2022], at a zoom level of \[level–more details about resolution?\] and subtracting those numbers from the vultures' tag-mounted altimeter readings. Finally, since some tags were occasionally set to a much higher fix rate (e.g. every 1 minute) for short periods of time, we downsampled the data to a 10 minute fix rate to make sure that all individuals had comparable fix rates.

@@

For each focal individual, we characterized a pattern-based spatial phenotype (Webber et al. 2023) by calculating a variety of movement and space use measures over the course of each of the eight seasons. We calculated home range as the 95% AKDE of a vulture's GPS fixes in a season @

>>>>

#### Social network centrality is known to be important in animals.

@@

#### We observe that some individuals are more central than others in a social network. What causes this variation?

An individual's social centrality may be determined by many factors. Differences in social position have been attributed to individual phenotypic characteristics, such as age, sex, or personality \[REFS\].

##### Individual traits

-   Multiple spatial behaviors → Social centrality @albery2021

-   Exploration → Weak associations with more conspecifics, more movement between flocks. @aplin2013

-   Slow-exploring → More stable relationships @aplin2013

-   Personality → Social structure @aplin2013

-   Stress → Weaker parental bonds, in captivity @boogert2014

-   Stress → Foraging with more others, less choosy, more central, in captivity @boogert2014

-   Stress → Foraging with more others, less choosy, more central, in the wild @brandl2019

-   More exploratory → higher degree, in finches @moyers2018

-   Quick explorers → More interactions @moyers2018

-   Slower explorers → less central ("Yet, slower explorers overall did not travel shorter distances than faster explorers, indicating that a less central social network position was not merely a consequence of lower activity.") @snijders2014

-   Females → higher strength, more preferred partners (only during lactation) @wey2013

-   Females → more stable partnerships @wey2013

##### Individuals' spatial phenotypes are important determinants of their social centrality

Of course, how an animal uses space is also an important driver of how many conspecifics it interacts with, and how frequently. The relationship between movement, space use, and social interaction patterns has been an area of active research in recent years. Webber et al [-@webber2023] have characterized the variation in patterns of animal movement and space use as the "spatial phenotype," as part of a larger framework for understanding the relationships between social and spatial drivers of animal behavior. In this light, the "spatial phenotype" may be considered in addition to more conventional phenotypic measures to explain why some individuals are more central than others in their social networks.

-   Definition of "spatial phenotype" @webber2023

-   Definition of "spatial behaviour" @albery2021

-   Note that explicit spatial context is also important, and we didn't consider it. (We should have! How do we justify not doing that?)

    -   An individual's observed spatial phenotype results from a combination of its internal tendency toward movement and the external environmental conditions through which it moves [(Nathan et al. 2008)](https://www.zotero.org/google-docs/?IyWvRB)

##### Social interactions occur when individuals encounter each other, in various behavioral situations.

-   Co-occurrence in space and time is necessary for social interaction to occur. Therefore, more movement → more social interaction, in general.

-   However, movement is also an expression of an individual's underlying tendency to move, as well as potentially an emergent consequence of things like dominance rank or social environment.

-   Social centrality arises from a combination of the number of interactions (strength) and unique individuals encountered. An individual can achieve high social centrality by being more geographically central in the population, versus by being more versatile.

-   No one has investigated the link between spatial phenotype and social centrality. We know that individuals vary in both spatial phenotype and social centrality, but we don't know how the one influences the other.

    -   Furthermore, social networks are not monolithic. Because the biological mechanisms behind social network structure are different in different social situations, it stands to reason that spatial phenotype should be related to social centrality in different ways in the different social contexts.

        -   For example, different components of spatial phenotype might be expected to relate differently to networks in different social situations.

        -   Also, we might just see different-direction effects. For example, individuals that move a lot might interact more frequently in flight but interact less frequently at feeding stations (either because they spend less time feeding or because they feed at more peripheral locations). Therefore, it is important to consider different behavioral situations separately when teasing apart the relationship between the social and spatial phenotypes.

@@

##### We set out to uncover whether spatial phenotypes could explain observed variation in social centrality \[should we add centroid location as a factor?\]

-   Hypotheses/predictions from slides

##### Research Questions

@@

### Study system

-   Unusual GPS tracking coverage of the population--high proportion of individuals tracked, so we can get a population-level sense of movement behavior and social interaction patterns

    -   Communal roosts: many individuals together at once, variability in size/density of roosts. Used for information exchange \[REFS\] and also important for disease transmission \[or not--cf Elvira Nili\]. Breeding/nesting also occurs at/near roost sites, on cliffs.

@@

## Field methods

@@

-   Download data from movebank (ornitela, inpa). Give general dates.

-   (Fix IDs)\--don't need to report this in the manuscript

-   Remove invalid periods in the hospital and capture cages (1 sentence)

    -   We subsetted the dataset to exclude periods of time that vultures spent in capture cages, in the wildlife hospital following poisonings or injuries, or in captivity prior to wild release (in the case of captive-born birds).

-   Data cleaning

    -   Basic data quality filters

        -   We cleaned the data to remove invalid GPS fixes.

    -   Remove spiky speeds

-   Double-masking process

-   Split into seasons

    -   Then we split data from each into three seasons. The fall, or "pre-breeding," season includes dates from 15 September through 14 December; the breeding season lasts from 15 December through 14 May of the following year; and the summer lasts from 15 May through 14 September. These season boundaries were chosen based on biological knowledge of the vultures' breeding timelines (see more details in the Study System section). Therefore, the analysis contained a total of 8 seasons, from the fall of 2020 through the 2023 breeding season (Table X).

-   Further cleaning for movement data

Then we split data from each into three seasons. The fall, or "pre-breeding," season includes dates from 15 September through 14 December; the breeding season lasts from 15 December through 14 May of the following year; and the summer lasts from 15 May through 14 September. These season boundaries were chosen based on biological knowledge of the vultures' breeding timelines (see more details in the Study System section). Therefore, the analysis contained a total of 8 seasons, from the fall of 2020 through the 2023 breeding season (Table X). All of these data was reserved for use in proximity-based social network analysis. Since the GPS tags give few to no fixes overnight, and since reception is often very poor on the cliffs where vultures roost, we combined information from individuals' last locations of the evening and first locations of the morning to assign a roost site for each individual on each night \[more details\]. Full details about the data cleaning and roost assignment procedure can be found in the supplementary material.

@@

For each season, we calculated movement measures only for individuals tracked for at least 30 days (after removing low-point days as described above). We calculated flight altitude by extracting ground elevation information from elevation rasters obtained from the `elevatr` R package [@hollister2022], at a zoom level of \[level--more details about resolution?\] and subtracting those numbers from the vultures' tag-mounted altimeter readings. Finally, since some tags were occasionally set to a much higher fix rate (e.g. every 1 minute) for short periods of time, we downsampled the data to a 10 minute fix rate to make sure that all individuals had comparable fix rates.

@@

For each focal individual, we characterized a pattern-based spatial phenotype (Webber et al. 2023) by calculating a variety of movement and space use measures over the course of each of the eight seasons. We calculated home range as the 95% AKDE of a vulture's GPS fixes in a season \@
